A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-07-25 tidyindex.qmd
proofread the GGGI example section

Indexes are commonly-used tools to monitor, communicate and make decisions across various domains in society. Their applications can be found in the environment (e.g. Air Quality Index, El Niño-Southern Oscillation Index), in the economy (Consumer Price Index, stock market indexes), and in the social sciences (e.g. QS University Rankings, Human Development Index).

To construct an index, experts typically first define an underlying concept of interest to measure, then collect relevant variables for measuring the concept, and lastly summarise these variables into a single numerical or categorical value. An index can be as simple as counting the number of days when the maximum temperature is greater than 25 degree. This is "the number of summer days" (SU), a climate index listed in ETCCDI Climate Change Indices [@climate-indexes]. In some cases, indexes are weighted linear combinations of variables. For example, Human Development Index (HDI) [@hdi] combines the three aspects (life expectancy, years of schooling, and GNI per capita) to measure country's development. In more complex scenarios, indexes are constructed by combining multiple variables, often accompanied by spatial and temporal considerations arising from diverse data formats, including in-situ stations, satellite imageries, and model outputs. An example of this is the Combined Drought Index (CDI) [@cdi], curated by the European Drought Observatory, which uses a decision tree to combine Standardized Precipitation Index (SPI) [@spi] with the satellite measure, Fraction of Absorbed Photosynthetically Active Radiation (FAPAR), and modelled soil moisture.

Our initial interest is to understand how different variable combinations contribute to the uncertainty of an index. However, as we start to understand how indexes are composed, we find many other steps in the construction also introduce uncertainty. In order to evaluate the uncertainty, an index needs to be broken down into its fundamental building blocks to analyse the effect of each component. This shifts the focus to investigate the decomposition of indexes into their constituent parts for analysing uncertainty. While the OECD handbook [@oecd_handbook_2008] has provided a set of steps for constructing composite socio-economic indexes, there is a need to extend these guidelines to accommodate more complex construction steps required for indexes in general.

@@

@@

@@

```{r idx-tour, fig.align='center', fig.height= 12, fig.width=12}
#| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index. The weights vary slightly from the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full animation is available at [vimeo link]."
frames <- c("001", "013", "037", "061", "085")
ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl, ncol = 3) 
```

>>>>

Indexes are commonly used to combine multivariate information into a single number for monitoring, communicating, and decision-making across various domains in society. Their applications can be found in the environment (e.g. Air Quality Index, El Niño-Southern Oscillation Index), in the economy (Consumer Price Index, stock market indexes), and in the social sciences (e.g. QS University Rankings, Human Development Index).

To construct an index, experts typically first define an underlying concept of interest to measure. This concept is latent and often without direct measure but of social and public importance. Relevant and available variables are then defined and collected for measuring the concept and combined as an index using statistical methods. Many decisions are factored into the construction process of an index. For example, decision on which variables to include may depend on the data availability and the choice in other similar indexes. Indexes using weighted linear combination of variables, for example, Human Development Index (HDI) [@hdi] combines the three aspects (life expectancy, years of schooling, and GNI per capita), needs to decide on the weights of each variable. In the case of a drought index, analysts may need to decide among similar indexes, on the time scale to aggregate precipitation, and the distribution used to fit the data. All these decisions can introduce uncertainty to the index and lead to different interpretation and actions in operation.

To understand how different factors contribute to the uncertainty of an index, it needs to be broken down into its fundamental building blocks to analyse the effect of each component. Indexes from various disciplines, i.e. [...] seems very different, can be boiled down into a [...]. Such decomposition of indexes standardise the index construction pipeline and offers benefit for comparing among indexes. In social indexes, the OECD handbook [@oecd_handbook_2008] has provided a set of steps for constructing composite socio-economic indexes, but there is a need to extend these guidelines to accommodate more complex construction steps required for indexes in general.

@@

The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023]. The index is composed of 14 variables, expressed as female-to-male ratios, which are first aggregated in a linear combination into the four dimensions using the weight from the `V-weight` column in @tbl-gggi-weights. The weight is calculated as the inverse of the standard deviation of each variable and scaling to sum to 1 within each dimension to allow a one percentage point change in the standard deviation of each variable to contribute equally to the index. The four dimensions are then aggregated in another linear combination with equal weight tp obtain the index. The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi`, along with the corresponding weights `gggi_weights`.

```{r tbl-gggi-weights}
#| tbl-cap: Weights of the fourteen variables in Global Gender Gap Index
#| tbl-cap-location: bottom
gggi_weights %>%  
   mutate(variable = str_replace_all(variable, "_", " ") %>%
            str_to_sentence()) %>% 
  select(c(1,2, 5:7)) %>% 
  mutate(dimension = c(rep("Economy", 5), rep("Education" ,4), 
                       rep("Health", 2), rep("Politics", 3))) %>% 
  knitr::kable(
    digits = 3, 
    col.names = c("Variable", "Dimension", "V-weight",
        "D-weight", "Weight")
    ) 
```

A natural thing to do when provided with the index data is to reproduce the index. This helps index analysts to verify the index calculation and become familiar with the methodology. For GGGI, the construction can be simplified as a single linear aggregation step in the dimension reduction module, with the `Weight` column in @tbl-gggi-weights, which is the product of the variable weight (`V-weight`) and dimension weight (`D-weight`).

````
gggi %>% 
  init(id = country) %>%
  add_meta(gggi_weights, var_col = variable) %>% 
  dimension_reduction(
    index_new = aggregate_linear(
      ~labour_force_participation:years_with_female_head_of_state,
      weight = weight)) 
````

The result can be compared with the GGGI values available in the report as shown in @fig-compare-gggi, validating the reproducibility of the index for country with no missing variables.

@@

To understand the uncertainty of this dimension reduction step while avoiding the missingness issue on the variable level, we can run a local tour to slightly vary the weight of each dimension to see how index value and country ranking changes. We select countries in the South Asia and Sub-Saharan Africa region and gradually increase the weight of one variable and reduce back to equal weight all four dimensions, one at a time, to produce an animation of how GGGI changes in each country. Five frames (equal weights and one for each dimension with a relatively higher weight) selected from the tour animation are shown in @fig-idx-tour and you can find the link to the full animation in figure caption.

Many insights can be derived from the animation and these selected frames. These can be helpful in understanding the sensitivity of the index value and country ranking to each variable and identifying the strengths and weaknesses of each country relative to others. When economy and education are given a greater weight, the variation in index values and ranking highlights countries that performs relatively better (moving right) or worse (moving left) in these two single dimensions. For example, the index moves to the left for Ethiopia and Senegal when economy is given a higher weight. This reveals the economy as a weakness for these two countries compared to their similarly-ranked peers. When health receives a higher weight, the impact on the index value is minimal. Notably, increasing the weight in politics leads to a pronounced decrease of index value for almost all the countries. Given the index value has a direct interpretation on how much of the gender gap has been closed, such a weight variation can cast a negative light on the progression to gender parity. Analysts need to further investigate the construction of the politics dimension and consider whether this is desirable for the index.

@@

```{r fig-idx-tour, fig.align='center', fig.height= 12, fig.width=12}
#| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index. The weights vary slightly from the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full animation is available at https://vimeo.com/847874016?share=copy."
frames <- c("001", "013", "037", "061", "085")
ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl, ncol = 3) 
```

### 2023-07-25 tidyindex.tex
proofread the GGGI example section

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Indexes are commonly-used tools to monitor, communicate and make
decisions across various domains in society. Their applications can be
found in the environment (e.g.~Air Quality Index, El Niño-Southern
Oscillation Index), in the economy (Consumer Price Index, stock market
indexes), and in the social sciences (e.g.~QS University Rankings, Human
Development Index).

To construct an index, experts typically first define an underlying
concept of interest to measure, then collect relevant variables for
measuring the concept, and lastly summarise these variables into a
single numerical or categorical value. An index can be as simple as
counting the number of days when the maximum temperature is greater than
25 degree. This is ``the number of summer days'' (SU), a climate index
listed in ETCCDI Climate Change Indices (Zhang 2020). In some cases,
indexes are weighted linear combinations of variables. For example,
Human Development Index (HDI) (United Nations Development Programme
2022) combines the three aspects (life expectancy, years of schooling,
and GNI per capita) to measure country's development. In more complex
scenarios, indexes are constructed by combining multiple variables,
often accompanied by spatial and temporal considerations arising from
diverse data formats, including in-situ stations, satellite imageries,
and model outputs. An example of this is the Combined Drought Index
(CDI) (Sepulcre-Canto et al. 2012), curated by the European Drought
Observatory, which uses a decision tree to combine Standardized
Precipitation Index (SPI) (McKee, Doesken, and Kleist 1993) with the
satellite measure, Fraction of Absorbed Photosynthetically Active
Radiation (FAPAR), and modelled soil moisture.

Our initial interest is to understand how different variable
combinations contribute to the uncertainty of an index. However, as we
start to understand how indexes are composed, we find many other steps
in the construction also introduce uncertainty. In order to evaluate the
uncertainty, an index needs to be broken down into its fundamental
building blocks to analyse the effect of each component. This shifts the
focus to investigate the decomposition of indexes into their constituent
parts for analysing uncertainty. While the OECD handbook (OECD, European
Union, and Joint Research Centre - European Commission 2008) has
provided a set of steps for constructing composite socio-economic
indexes, there is a need to extend these guidelines to accommodate more
complex construction steps required for indexes in general.

@@

\includegraphics{tidyindex_files/figure-pdf/gggi-verify-1.pdf}

@@

@@

\caption{Five frames selected from varying the linear weights of four
dimensions in Global Gender Gap Index. The weights vary slightly from
the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe
how the index and ranking reponse. Full animation is available at
{[}vimeo link{]}.}

@@

\leavevmode\vadjust pre{\hypertarget{ref-cdi}{}}%
Sepulcre-Canto, G., S. Horion, A. Singleton, H. Carrao, and J. Vogt.
2012. {``Development of a {Combined} {Drought} {Indicator} to Detect
Agricultural Drought in {Europe}.''} \emph{Natural Hazards and Earth
System Sciences} 12 (11): 3519--31.
\url{https://doi.org/10.5194/nhess-12-3519-2012}.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

Indexes are commonly used to combine multivariate information into a
single number for monitoring, communicating, and decision-making across
various domains in society. Their applications can be found in the
environment (e.g.~Air Quality Index, El Niño-Southern Oscillation
Index), in the economy (Consumer Price Index, stock market indexes), and
in the social sciences (e.g.~QS University Rankings, Human Development
Index).

To construct an index, experts typically first define an underlying
concept of interest to measure. This concept is latent and often without
direct measure but of social and public importance. Relevant and
available variables are then defined and collected for measuring the
concept and combined as an index using statistical methods. Many
decisions are factored into the construction process of an index. For
example, decision on which variables to include may depend on the data
availability and the choice in other similar indexes. Indexes using
weighted linear combination of variables, for example, Human Development
Index (HDI) (United Nations Development Programme 2022) combines the
three aspects (life expectancy, years of schooling, and GNI per capita),
needs to decide on the weights of each variable. In the case of a
drought index, analysts may need to decide among similar indexes, on the
time scale to aggregate precipitation, and the distribution used to fit
the data. All these decisions can introduce uncertainty to the index and
lead to different interpretation and actions in operation.

To understand how different factors contribute to the uncertainty of an
index, it needs to be broken down into its fundamental building blocks
to analyse the effect of each component. Indexes from various
disciplines, i.e.~{[}\ldots{]} seems very different, can be boiled down
into a {[}\ldots{]}. Such decomposition of indexes standardise the index
construction pipeline and offers benefit for comparing among indexes. In
social indexes, the OECD handbook (OECD, European Union, and Joint
Research Centre - European Commission 2008) has provided a set of steps
for constructing composite socio-economic indexes, but there is a need
to extend these guidelines to accommodate more complex construction
steps required for indexes in general.

@@

The Global Gender Gap Index (GGGI), published annually by the World
Economic Forum, measures gender parity by assessing relative gaps
between men and women in four key areas: Economic Participation and
Opportunity, Educational Attainment, Health and Survival, and Political
Empowerment (World Economic Forum 2023). The index is composed of 14
variables, expressed as female-to-male ratios, which are first
aggregated in a linear combination into the four dimensions using the
weight from the \texttt{V-weight} column in
Table~\ref{tbl-gggi-weights}. The weight is calculated as the inverse of
the standard deviation of each variable and scaling to sum to 1 within
each dimension to allow a one percentage point change in the standard
deviation of each variable to contribute equally to the index. The four
dimensions are then aggregated in another linear combination with equal
weight tp obtain the index. The 2023 GGGI data is available from the
Global Gender Gap Report 2023 in the country's economy profile and can
be accessed in R via the \texttt{tidyindex} package as \texttt{gggi},
along with the corresponding weights \texttt{gggi\_weights}.

\hypertarget{tbl-gggi-weights}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.5455}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1299}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0909}}@{}}
\caption{\label{tbl-gggi-weights}Weights of the fourteen variables in
Global Gender Gap Index}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
V-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
D-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Weight
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
V-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
D-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Weight
\end{minipage} \\
\midrule()
\endhead
Labour force participation & Economy & 0.199 & 0.25 & 0.050 \\
Wage equality for similar work & Economy & 0.310 & 0.25 & 0.078 \\
Estimated earned income & Economy & 0.221 & 0.25 & 0.055 \\
Legislators senior officials and managers & Economy & 0.149 & 0.25 &
0.037 \\
Professional and technical workers & Economy & 0.121 & 0.25 & 0.030 \\
Literacy rate & Education & 0.191 & 0.25 & 0.048 \\
Enrolment in primary education & Education & 0.459 & 0.25 & 0.115 \\
Enrolment in secondary education & Education & 0.230 & 0.25 & 0.058 \\
Enrolment in tertiary education & Education & 0.121 & 0.25 & 0.030 \\
Sex ratio at birth & Health & 0.693 & 0.25 & 0.173 \\
Healthy life expectancy & Health & 0.307 & 0.25 & 0.077 \\
Women in parliament & Politics & 0.310 & 0.25 & 0.078 \\
Women in ministerial positions & Politics & 0.247 & 0.25 & 0.062 \\
Years with female head of state & Politics & 0.443 & 0.25 & 0.111 \\
\bottomrule()
\end{longtable}

A natural thing to do when provided with the index data is to reproduce
the index. This helps index analysts to verify the index calculation and
become familiar with the methodology. For GGGI, the construction can be
simplified as a single linear aggregation step in the dimension
reduction module, with the \texttt{Weight} column in
Table~\ref{tbl-gggi-weights}, which is the product of the variable
weight (\texttt{V-weight}) and dimension weight (\texttt{D-weight}).

\begin{verbatim}
gggi %>% 
  init(id = country) %>%
  add_meta(gggi_weights, var_col = variable) %>% 
  dimension_reduction(
    index_new = aggregate_linear(
      ~labour_force_participation:years_with_female_head_of_state,
      weight = weight)) 
\end{verbatim}

The result can be compared with the GGGI values available in the report
as shown in Figure~\ref{fig-compare-gggi}, validating the
reproducibility of the index for country with no missing variables.

@@

To understand the uncertainty of this dimension reduction step while
avoiding the missingness issue on the variable level, we can run a local
tour to slightly vary the weight of each dimension to see how index
value and country ranking changes. We select countries in the South Asia
and Sub-Saharan Africa region and gradually increase the weight of one
variable and reduce back to equal weight all four dimensions, one at a
time, to produce an animation of how GGGI changes in each country. Five
frames (equal weights and one for each dimension with a relatively
higher weight) selected from the tour animation are shown in
Figure~\ref{fig-idx-tour} and you can find the link to the full
animation in figure caption.

Many insights can be derived from the animation and these selected
frames. These can be helpful in understanding the sensitivity of the
index value and country ranking to each variable and identifying the
strengths and weaknesses of each country relative to others. When
economy and education are given a greater weight, the variation in index
values and ranking highlights countries that performs relatively better
(moving right) or worse (moving left) in these two single dimensions.
For example, the index moves to the left for Ethiopia and Senegal when
economy is given a higher weight. This reveals the economy as a weakness
for these two countries compared to their similarly-ranked peers. When
health receives a higher weight, the impact on the index value is
minimal. Notably, increasing the weight in politics leads to a
pronounced decrease of index value for almost all the countries. Given
the index value has a direct interpretation on how much of the gender
gap has been closed, such a weight variation can cast a negative light
on the progression to gender parity. Analysts need to further
investigate the construction of the politics dimension and consider
whether this is desirable for the index.

@@

\caption{\label{fig-idx-tour}Five frames selected from varying the
linear weights of four dimensions in Global Gender Gap Index. The
weights vary slightly from the official simple average weights (0.25,
0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full
animation is available at https://vimeo.com/847874016?share=copy.}

@@
